---
title: "Day 4: HIV Prevalence"
author: "Cleo Falvey, Filip Stefanovic, & Sydney Horan"
date: "3/7/2021"
output: html_document
---

## Making Graphs about Malaria Outbreaks Over Time

Good morning! Welcome to Day 4. Today, we'll be showing you how to make graphs about malaria outbreaks over time. But before, we do that, we need to load all necessary packages. The Tidyverse is always a good place to start.

```{r}

library(tidyverse) # for tidying
library(scales) # for renaming labels
library(gganimate) # for potential animation

```

### Reading In Data

Next, we download the data from Gapminder. You can choose any data set, but I wanted to choose a data set that was standardized by population (i.e., cases per 100k people.) When you do your project, you can choose either case counts or cases per population, but keep in mind that countries with larger populations (will likelyl) have higher case counts.

```{r}
# read  in dataset with read_csV()
malaria <- read_csv("malaria_cases_per_100000_reported.csv")

head(malaria)
```

Looking at the data, we can see we have case count data that's grouped by both country and year. This structure of data is fairly typical for epidemiological data. 

### Cleaning Data

The next step in the pathway is cleaning our data. What I'm doing right here is pivoting this data from "long" to "wide." You can see the difference once the _head()_ function is called. The _pivot_longer()_ function takes the columns that we are pivoting - notice I indexed them numerically (and that arrays in R start at 1). The next part of the pivoting argument simply helps us understand what we are renaming the variables to when we look at our new data set.

```{r}
# clean data
malaria_longer <- malaria %>% # data cleaning
  pivot_longer(cols = 2:18,
               names_to = "year",
               values_to = "cases") # long format because the years as a title of the columns was not helping because we wouldn't then be able to graph by year

head(malaria_longer)


```

### Manipulating Data 

```{r}
most_malaria <- malaria_longer %>%
  group_by(country) %>%
  summarize_if(is.numeric, mean, na.rm=T) %>%
  arrange(desc(cases))

```


Here are two data tables showing countries with the highest and least  prevalence of malaria on average across the time span selected (1990-2006). Now that we know what these countries are, we will graph these countries to see what trends emerge.

```{r}
head(most_malaria)

tail(most_malaria)


interesting_countries <- c("Zambia", "Malawi","Sao Tome and Principe","Tanzania","Solomon Islands","Algeria","Morocco","Turkmenistan","Russia","Uzbekistan")  # remember, we always put characters in quotes!

```

### Graphing Data

Now that we have our data manipulated and cleaned, let's graph and tell a story. This is the first graph I made. Do you see any issues with it? One of the biggest issues is that it's too much data to display - the scales are wildly different between the top five highest and lowest countries. Let's make two different ones, and place them side by side.

```{r}
malaria_longer %>% # take malaria dataset
  filter(country %in% interesting_countries) %>% #filter to interesting countries
  ggplot(aes(x=year, y=cases, color=country)) + # set up ggplot
  geom_point() + #add what geometry we want
  scale_x_discrete(breaks=pretty_breaks()) + # fix messy scales
  labs(title="Malaria Cases in the Top 5 Highest and Lowest Countries")
```



```{r}
lowest_countries <- c("Algeria","Morocco","Turkmenistan","Russia","Uzbekistan")  # remember, we always put characters in quotes!

highest_countries <- c("Zambia", "Malawi","Sao Tome and Principe","Tanzania","Solomon Islands")

# highest prevalence
malaria_longer %>% # take malaria dataset
  filter(country %in% highest_countries) %>% #filter to interesting countries
  ggplot(aes(x=year, y=cases, color=country)) + # set up ggplot
  geom_point() + #add what geometry we want
  scale_x_discrete(breaks=pretty_breaks()) + # fix messy scales
  labs(title="Malaria Cases in the Top 5 Highest Countries")


# lowest prevalence
malaria_longer %>% # take malaria dataset
  filter(country %in% lowest_countries) %>% #filter to interesting countries
  ggplot(aes(x=year, y=cases, color=country)) + # set up ggplot
  geom_point() + #add what geometry we want
  scale_x_discrete(breaks=pretty_breaks()) + # fix messy scales
  labs(title="Malaria Cases in the Top 5 Lowest Countries") +
  scale_color_brewer(palette="Set1")
  
```

### Synthesizing and Researching

Now that we've made our graphs, let's do some research. We could look at what the lowest prevalence countries have in common - is it a lack of mosquitoes? A public health campaign that successfully eradicated malaria? What do the highest prevalence countries have in common? An aged medical infrastructure? Lots and lots of mosquitoes? Here are a few journal articles (scientific, scholarly) which I think might be worth reading and citing if I were doing this project.

### Tips and Tricks

Remember, as you do more research, you may uncover more interesting things to graph! This is a fluid process and involves a lot of trial and error to figure out what's the best story to tell and what might make sense or not. 

# Miscellaneous Takeaways
- Research is a journey.
- Think about what makes a graph look good. Is it colorful (and colorblind-friendly)? Is it clear? Do the labels look good? Is the key clear?